---
title: "Disease Progression Models"
author: "Tyler Dunlap, PharmD"
execute:
  eval: false
format:
  html:
    title-block-banner: true
    self-contained: true
    toc: true
    toc-location: left
    css: styles.css
    theme: journal
    page-layout: article
editor: visual
bibliography: references.bib
---

# Disease Progression Models

```{r}
#| warning: false
#| error: false
#| message: false
library(tidyverse)
library(mrgsolve)
```

## Overview

Disease and pharmacodynamic modeling can enhance trial efficiency in discrimination of symptomatic and disease modifying effects. Disease progression models aim to capture the time course of disease, often over time scales of months to years. When disease progression is insidious, quantifying a treatment effect relative to placebo or standard of care must take into consideration the changing disease status of the patients over time. Examples of disease progression models are relatively rare compared to PK/PD models, but theyare essential to clinical trial simulation, thus worth spending some time with to really understand them. Importantly, placebo effects may be explicitly modeled so that the drug effect can be estimated as the observed effect minus the placebo effect. Additionally, longitudinal evaluation of drug effect may be to understand whether the drug actually modifies the time course of disease progression or simply provides symptomatic relief.

### Types of Disease Progression Models

Simulation from the disease model depends on the level of understanding of the disease process. Models for disease can be empirical, semimechanistic, mechanistic, or a combination of each. Many types of disease progression models are possible, but the most common used in pharmacometrics is mixed effects models of course. However, its worth knowing other types of models may be used.

| Model                | Description                                                                                                                                                                                                                                                                  |
|----------------|-------------------------------------------------------|
| Mixed effects models | These models can be applied to understand disease progression dynamics within populations and in individual patients.                                                                                                                                                        |
| Multistate models    | These models categorize individuals into different compartments or states, representing various disease stages and describe transition intensities of patients between states. There is an entire section of this book dedicated to multistate models. They are fascinating. |
| Markov Models        | These models use the Markov property, where future states only depend on the current state. They're used to study transitions between discrete disease stages.                                                                                                               |
| Stochastic Models    | These models incorporate randomness to account for random processes, uncertainty, and variability in disease progression. They're useful for capturing individual-level variation.                                                                                           |
| Mechanistic Models   | These models may be classified as "Systems Biology" or "Quantitative Systems Pharmacology" and are models that incorporate biological mechanisms at multiple levels of the system (i.e., organs, tissues, cells).                                                            |

## Treatment Effects

### Symptomatic

Symptomatic effects are drug effects that act directly on disease status. They may have an immediate or slow onset, but the key feature to understand is that when the drug is removed, the patient's disease status returns to the natural disease progression level as if they had not received drug at all.

### Disease-Modifying

Disease-modifying treatment effects alter the time-varying components of the disease model. In other words, they can actually slow the progression of disease by reducing the disease progression rate.

### Placebo Effects

Placebo effects typically have a quick onset and quick offset, or washout. They are modeled explicitly in disease progression models to estimate the absolute drug effect by capturing "observed" drug effects in untreated patients.

### Natural disease progression

Natural disease progression is typically assumed to follow a linear or nonlinear progression with time. Below is a linear regression model representing natural disease progression in which the disease status at time $t$ is a function of the patient's baseline disease status $Disease_0$ and the amount of time that has passed. The progression rate is parameterized with a $k_{prog}$ parameter which quantifies how quickly the disease progresses with time.

$$
Disease_t = Disease_0 + k_{prog}*t
$$

```{r}
#| warning: false
#| error: false
#| message: false
# natural disease progression model
ndp_code <- '
$PARAM @annotated
TVDIS0   : 0.6 : Baseline disease status
TVKPROG  :  0.01  : Disease progression rate

$OMEGA @annotated @block @correlation
eDIS0   : 0.09  : eta baseline disease status
eKPROG  : 0.3 0.09  : eta disease progression rate

$MAIN
double DIS0 = TVDIS0*exp(eDIS0);
double KPROG = TVKPROG*exp(eKPROG);
double DIS_T = DIS0 + KPROG*TIME;

$CAPTURE @annotated
DIS_T  : Disease status at time T
'
ndp <- mcode("ndp", ndp_code)
```

```{r}
#| warning: false
#| error: false
#| message: false
dp <- data.frame(ID=1:10)

ndp %>%
  data_set(dp) %>%
  mrgsim(end=100, delta=1) %>%
  plot(DIS_T~time)
```

### Placebo effects

$$
E_{placebo} = Am* (\frac{k_{on}}{k_{on}-k_{off}})(e^{-k_{off}*t}-e^{-k_{on}*t})
$$

```{r}
#| warning: false
#| error: false
#| message: false
# placebo effect model
placebo_code <- '
$PARAM @annotated
TVAM    : 20  : 
TVKON   : 0.6 : effect onset parameter
TVKOFF  :  0.1  : effect "dissociation" parameter

$OMEGA @annotated 
eAm   : 0.09  : eta baseline disease status
ekon  : 0.09  : eta effect onset
ekoff : 0.09  : eta effect dissociation (wearing off)

$MAIN
double Am = TVAM*exp(eAm);
double kon = TVKON*exp(ekon);
double koff = TVKOFF*exp(ekoff);
double P_effect = Am*(kon/(kon-koff))*(exp(-koff*TIME)-exp(-kon*TIME));

$CAPTURE @annotated
P_effect  : Placebo effect at time T
'

p_effect <- mcode("placebo_code", placebo_code)
```

```{r}
#| warning: false
#| error: false
#| message: false
p <- data.frame(ID=1:10)

p_effect %>%
  data_set(p) %>%
  mrgsim(end=100, delta=1) %>%
  #obsaug() %>%
  plot(P_effect~time)
```

### Symptomatic effects

$$
E_{sym} = C_p*k_s*(1-e^{-k_{es}*t})
$$

```{r}
#| warning: false
#| error: false
#| message: false
symptomatic_code <- '
$PARAM @annotated
TVCP   :  20    : average plasma concentration
TVKS   :  0.6   : effect scalar
TVKES  :  0.1   : effect wearing off 

$OMEGA @annotated 
eCP   : 0.09  : eta baseline disease status
eKS  : 0.09  : eta effect onset
eKES : 0.09  : eta wearing off

$MAIN
double CP = TVCP*exp(eCP);
double KS = TVKS*exp(eKS);
double KES = TVKES*exp(eKES);
double S_EFF = CP*KS*(1 - exp(-KES*TIME));

$CAPTURE @annotated
S_EFF  : Symptomatic effect at time T
'

s_effect <- mcode("symptomatic_code", symptomatic_code)
```

```{r}
#| warning: false
#| error: false
#| message: false
s <- data.frame(ID=1:10)

s_effect %>%
  data_set(p) %>%
  mrgsim(end=100, delta=1) %>%
  plot(S_EFF~time)
```

### Disease modifying effects

$$
E_{dm}=C_p*k_m
$$

```{r}
#| warning: false
#| error: false
#| message: false
dm_code <- '
$PARAM @annotated
TVCP   :  20    : average plasma concentration
TVKM   :  0.6   : Modifying effect

$OMEGA @annotated 
eCP   : 0.09  : eta average plasma concentration
eKM   : 0.09  : eta modifying effect

$MAIN
double CP = TVCP*exp(eCP);
double KM = TVKM*exp(eKM);
double DM_EFF = CP*KM;

$CAPTURE @annotated
DM_EFF  : Disease modifying effect
'

dm_effect <- mcode("dm_code", dm_code)
```

```{r}
#| warning: false
#| error: false
#| message: false
dm <- data.frame(ID=1:10)

dm_effect %>%
  data_set(dm) %>%
  mrgsim(end=100, delta=1) %>%
  plot(DM_EFF~time)
```

### Drug effects (Drug-Placebo)

$$
Dis_{drug} = Dis_0+(k_{prog}-C_p*k_m)*t - placebo - C_p*k_s*(1-e^{-k_{es}*t})
$$

```{r}
#| warning: false
#| error: false
#| message: false
drug_effect <- '
$PARAM @annotated
// Pharmacokinetic Model
TVKA   : 0.6 : Absorption rate constant (1/hr)
TVCL   :  6  : Clearance (volume/time)
TVV1    : 15  : Central volume (volume)

// Disease Progression model
TVDIS0 : 1   : Baseline disease status
TVKPROG : 1  : Linear disease progression rate

// Disease Modifying effect
TVKM    : 0.1 : Disease modifying effect of drug

// Symptomatic effect
TVKS    : 1     : Symptomatic effect
TVKEO   : 0.1   : Wearing off of symptomatic effect

// Placebo Effect
TVAM    : 20    : 
TVKON   : 0.6   : effect onset parameter
TVKOFF  :  0.1  : effect "dissociation" parameter

$PARAM @covariates @annotated
PLACEBO : 1   : PLACEBO
TRT     : 1   : TREATMENT

$CMT  @annotated
ABS    : Extravascular compartment (mass)
CENT   : Central compartment (mass)

$OMEGA @annotated
// PK parameters
eKA : 0.09  : eta KA
eCL : 0.09  : eta CL
eV1  : 0.09  : eta V

// Disease progression
eDIS0 : 0.09 : eta baseline disease status
eKPROG : 0.09 : eta disease progression rate

// Disease modifying effect
eKM   : 0.09  : eta disease modifying effect

// symptomatic effect
eKS   : 0     : eta symptomatic effect
eKEO  : 0     : eta wear off of symptomatic effect

// Placebo effect
eAm   : 0.09  : eta baseline disease status
ekon  : 0.09  : eta effect onset
ekoff : 0.09  : eta effect dissociation (wearing off)

$GLOBAL
#define CP (CENT/V1)

$MAIN
// PK parameters
double KA = TVKA*exp(eKA);
double CL = TVCL*exp(eCL);
double V1 = TVV1*exp(eV1);
double DIS0 = TVDIS0*exp(eDIS0);
double KPROG = TVKPROG*exp(eKPROG);
double KM = TVKM*exp(eKM);
double KS = TVKS*exp(eKS);
double KEO = TVKEO*exp(eKEO);
double Am = TVAM*exp(eAm);
double kon = TVKON*exp(ekon);
double koff = TVKOFF*exp(ekoff);

// Modifying effect
double DM_EFF = CP*KM;

// Symptomatic effect
double S_EFF = CP*KS*(1 - exp(-KEO*TIME));

// Placebo effect
double P_EFF = Am*(kon/(kon-koff))*(exp(-koff*TIME)-exp(-kon*TIME));

// Disease Status
double DISEASE = DIS0 + (KPROG - TRT*DM_EFF)*TIME - PLACEBO*P_EFF - TRT*S_EFF;

$ODE
dxdt_ABS    = -KA*ABS;
dxdt_CENT   =  KA*ABS - CL*CP;

$CAPTURE @annotated
PLACEBO : PLACEBO
TRT    : TRT
DISEASE : Disease status
CP : Plasma concentration (mass/volume)
'

drug_effect <- mcode("drug_effect", drug_effect)
```

```{r}
#| warning: false
#| error: false
#| message: false
de <- expand.ev(ID=1:10, amt=100, PLACEBO=c(0,1), TRT=c(0, 1))

drug_effect %>%
  data_set(de) %>%
  mrgsim(end=100, delta=1) %>%
  plot(DISEASE~time)
```

### Summary

A starting point for utilization of Disease - Progression Models is consideration of how baseline conditions change with time. Several basic patterns are known. Diseases can cause structural or functional alterations of all components of PK/PD models which requires extra care in evaluating data. Read [@mould2012] (and watch her youtube lecture from the NIH Principles of Clinical Pharmacology course).

### References
