---
title: "Mixture Models"
author: "Tyler Dunlap, PharmD"
execute:
  eval: false
format:
  html:
    title-block-banner: true
    self-contained: true
    toc: true
    toc-location: left
    css: styles.css
    theme: journal
    page-layout: article
editor: visual
bibliography: references.bib
---

## Mixture Models

```{r}
#| warning: false
#| error: false
#| message: false
library(tidyverse)
library(mrgsolve)
```

## Overview

Mixture models are a class of statistical models used for modeling data that can be thought of as a combination of multiple underlying subpopulations or distributions. In other words, mixture models are used to represent the idea that the observed data comes from a mixture of different sources or components, each with its own probability distribution. **Components or Clusters:** These are the underlying distributions that make up the mixture. Each component represents a distinct subpopulation or group within the data. For example, in a Gaussian mixture model, each component is a Gaussian (normal) distribution. **Component Weights:** These are the probabilities associated with each component. They represent the proportion of the data that comes from each component. The weights must sum up to 1.

### Types

| Mixture                       | Description                                                                                                                                               |
|-------------------|-----------------------------------------------------|
| **Gaussian Mixture Model**    | Each component is a Gaussian distribution (i.e., ETAs). These types of models are widely used for clustering and density estimation with continuous data. |
| **Multinomial Mixture Model** | Used for discrete data, where each component follows a multinomial distribution.                                                                          |

Here is a pharmacokinetic example demonstrating a mixture model. Say we have a drug in which the clearance is strongly associated with a mutation in a CYP3A4 drug-metabolizing enzyme. We haven't done any pharmacogenomic testing, but we may be able to infer the probability of the patient having the relevant mutation based on their observed drug clearance.

```{r}
#| warning: false
#| error: false
#| message: false
pkmix <- '
$PROB PK mixture model

$PLUGIN Rcpp

$PARAM @annotated
TVCL  : 10  :
TVV1  :  42  :
MUTEFF : -1.5  : effect of mutation

$PARAM @annotated @covariate
MUT   : 0    : mutation

$CMT CENT

$OMEGA @annotated
eCL  : 0.09  : eta clearance

$MAIN
if(NEWIND <=1) {
  int POP = R::rbinom(1,0.2);
}

double CL = TVCL*exp(eCL);
double V1 = TVV1;

if (POP==1) CL = TVCL*exp(MUTEFF*POP)*exp(eCL);

$ODE
dxdt_CENT = -(CL/V1)*CENT;

$TABLE
capture IPRED = CENT/V1;

$CAPTURE POP
'
  
pkmix <- mcode('pkmix', pkmix)
```

```{r}
#| warning: false
#| error: false
#| message: false
pkmix %>%
  ev(data.frame(ID=1:30, amt=250, cmt=1, time=0)) %>%
  mrgsim(end = 24, delta=1) %>%
  plot(IPRED~time|factor(POP))
```

The mixture may be composed of more than one population... **NOT DONE**

```{r}
#| warning: false
#| error: false
#| message: false
code <- '
$PROB Multiple populations mixture model

$PLUGIN Rcpp

$PARAM @annotated
TVBS    ;      ; baseline size
TVKG    :      : growth rate
TVKD    :      : death rate (drug effect)
TVF     :       :
P1      : 0.33 : 
P2      :       :
P3      :        :


$MAIN
if(NEWIND <=1) {
  double mix = R::runif(0,1);
  int POP = 1;
  if(mixv > p1) POP = 2;
  if(mixv > (p1+p2)) POP = 3;
}

$CAPTURE POP mixv
'
```

::: callout-note
Mixture models form the basis for some more complicated models such as "Cure" and "Hidden Markov Models" which are discussed in the "Frailty & Cure Models" and "Multistate Models" sections, respectively.
:::

### Summary

Mixture models are often used for various applications, including clustering, density estimation, anomaly detection, and more. They are especially useful when the data is a blend of multiple underlying processes, and traditional single-distribution models like Gaussians might not capture the complexity adequately. The parameters of a mixture model are typically estimated using techniques like the Expectation-Maximization (EM) algorithm or Variational Inference.

### References
