---
title: "Pharmacokinetic Models"
author: "Tyler Dunlap, PharmD"
execute:
  eval: false
format:
  html:
    title-block-banner: true
    self-contained: true
    toc: true
    toc-location: left
    css: styles.css
    theme: journal
    page-layout: article
editor: visual
---

# Pharmacokinetic Models

```{r}
#| warning: false
#| message: false
#| error: false
library(tidyverse)
library(mrgsolve)
library(DiagrammeR)
```

## Overview

Pharmacokinetics (PK) is the study of how drugs move through the body, or more colloquially, "what the body does to the drug", encompassing processes such as absorption, distribution, metabolism, and elimination (ADME). A PK model is a mathematical framework that can help quantify the time course of these processes to predict drug concentrations in the body over time, identify patient factors affecting drug disposition, facilitate implementation of dose optimization strategies, and much more. It may be my biased opinion, but these modeling endeavors are critical to promoting the rationale use of therapeutics and facilitating efficient drug development.

### Pharmacokinetic parameters

Below is a table of PK parameters commonly used in PK models. It is by no means comprehensive.

| Parameter                        | Description                                                                                                                           |
|--------------------|----------------------------------------------------|
| Clearance                        | The rate at which a drug is removed from the body (elimination or metabolism) relative to its concentration.                          |
| Volume of Distribution (Vd)      | The apparent space the drug appears to occupy in the body relative to its concentration.                                              |
| Bioavailability (F)              | The fraction of a drug that reaches systemic circulation after administration (oral, intravenous, etc.).                              |
| Absorption rate (Ka)             | A first order rate constant that is used to characterize the rate of "appearance" of drug into the plasma (or the relevant biomatrix) |
| lag time (TLAG)                  | The delay in time to apparent initiation of absorption (i.e., delayed release formulations)                                           |
| Cmax                             | Maximum concentration observed during a dosing window                                                                                 |
| Area under the curve (AUC)       | Area under the concentration-time curve; provides a more robust way to quantify drug exposure                                         |
| Intercompartmental clearance (Q) | Distribution rate between compartments (i.e., central and peripheral)                                                                 |

## Compartmental models

::: callout-important
An understanding of the difference between zero-, first-, and (rarely) second-order processes is absolutely crucial to working with compartmental models.
:::

#### One-compartment model

One-compartment models are the simplest compartmental models. Notably, when drugs are given orally, we include an "absorption compartment" but do not include this when we characterize how many "compartments" this drug follows because we only care about the rate of absorption from the absorption compartment, not the concentration of drug within it. The example below demonstrates a one-compartment model with first order elimination and first-order absorption.

```{r}
#| warning: false
#| message: false
#| error: false
pk1_code <- '
$PARAM @annotated
TVKA   : 0.6 : Absorption rate constant (1/hr)
TVCL   :  6  : Clearance (volume/time)
TVV1    : 15  : Central volume (volume)

$CMT  @annotated
ABS    : Extravascular compartment (mass)
CENT   : Central compartment (mass)

$OMEGA @annotated
eKA : 0.09  : eta KA
eCL : 0.09  : eta CL
eV1  : 0.09  : eta V

$GLOBAL
#define CP (CENT/V1)

$MAIN
double KA = TVKA*exp(eKA);
double CL = TVCL*exp(eCL);
double V1 = TVV1*exp(eV1);

$ODE
dxdt_ABS    = -KA*ABS;
dxdt_CENT   =  KA*ABS - CL*CP;

$SIGMA @annotated
PROP  : 0.05   : proportional error

$TABLE
capture DV = CP*(1+PROP);

// this resimulates observation if DV is negative
while(DV <= 0) {
  simeps();
  DV = CP*(1+PROP);
}

$CAPTURE @annotated
CP : Plasma concentration (mass/volume)
DV : dependent variable
TVKA : typical value first order absorption
TVCL : typical value clearance
TVV1 : typical value volume of distribution
'
  
pk1 <- mcode("pk1", pk1_code)
```

Place the dose into the "first" compartment to simulate intravenous dosing. In the code below, the zero_re() function is used to remove the random effects from the model. Comment this out with to see the interindividual variability (IIV) among the five simulated individuals. Use the addl argument to specify the number of additional doses and ii to specify the interdose interval (i.e., q12h, q24h). cmt=1 puts the dose into the absorption compartment.

```{r}
#| warning: false
#| message: false
#| error: false
e <- data.frame(ID=1:5, amt=100, ii=24, addl=3, cmt=1, evid=1)

pk1 %>%
  ev(amt=100) %>%
  zero_re() %>% # this removes random effects; comment out to see IIV
  mrgsim(end = 100, delta=1) %>% 
  plot(CP~time)
```

Place the dose into the "second" (cmt=2) compartment to simulate intravenous dosing. Modify ii and addl to change the inter-dose interval and the number of additional doses given, respectively. Change end and delta to modify the simulation length and observation frequency, respectively.

```{r}
#| warning: false
#| message: false
#| error: false
e <- data.frame(ID=1:5, amt=100, ii=24, addl=3, cmt=2, evid=1)

pk1 %>%
  ev(amt=100) %>%
  mrgsim(end = 100, delta=1) %>% 
  plot(CP~time)
```

#### Two-compartment model

In two-compartment models, the drug distributes between "central" and "peripheral" compartments. The central compartment is typically thought of as tissues well-perfused by the vascular supply (e.g., kidney) while peripheral compartments are thought of as tissues in which the drug distributes more slowly (e.g., bone).

```{r}
#| warning: false
#| message: false
#| error: false
pk2_code <- '
$PARAM @annotated
TVKA   : 0.6 : Absorption rate constant (1/time)
TVCL   :  6  : Clearance (volume/time)
TVV1    : 15  : Central volume (volume)
TVV2   : 15  : Peripheral volume (volume)
TVQ    : 5   : Intercompartment clearance (volume/time)

$CMT  @annotated
ABS    : Extravascular compartment (mass)
CENT   : Central compartment (mass)
PER    : Peripheral compartment (mass)

$OMEGA @annotated
eKA : 0.09  : eta KA
eCL : 0.09  : eta CL
eV1  : 0.09  : eta V
eV2 : 0.09  : eta V2
eQ  : 0.09  : eta Q

$GLOBAL
#define CP (CENT/V)
#define CT (PER/V2)

$MAIN
double KA = TVKA*exp(eKA);
double V1 = TVV1*exp(eV);
double CL = TVCL*exp(eCL);
double V2 = TVV2*exp(eV2);
double Q = TVQ*exp(eQ);

$ODE
dxdt_ABS    = -KA*ABS;
dxdt_CENT   =  KA*ABS - CL*CP - Q*CP + Q*CT;
dxdt_PER    =  Q*CP - Q*CT;

$SIGMA @annotated
PROP  : 0.05   : proportional error

$TABLE
capture IPRED = CENT/V;
capture DV = IPRED*(1+PROP);

while(DV <= 0) {
  simeps();
  DV = IPRED*(1+PROP);
}

$CAPTURE @annotated
CP : Plasma concentration (mass/volume)
TVKA   :  Absorption rate constant (1/time)
TVCL   : Clearance (volume/time)
TVV1    : Central volume (volume)
TVV2   : Peripheral volume (volume)
TVQ    : Intercompartment clearance (volume/time)
'
  
pk2 <- mcode("pk2", pk2_code)
```

Because I have included TVKA, TVCL, etc in the \$CAPTURE block, you can modify these parameters directly in the e object below.

```{r}
#| warning: false
#| message: false
#| error: false
e <- data.frame(ID=1:5, amt=100, ii=24, addl=3, cmt=1, evid=1, TVCL=8)

pk2 %>%
  ev(amt=100) %>%
  mrgsim(end = 100, delta=1) %>% 
  plot(CP~time)
```

#### Three-compartment model

The three-compartment model adds an additional peripheral compartment. Note that the peripheral compartments are both connected only to the central compartment, not to each other.

```{r}
#| warning: false
#| message: false
#| error: false
pk3_code <- '
$PARAM @annotated
TVKA   : 0.6 : Absorption rate constant (1/time)
TVCL   :  6  : Clearance (volume/time)
TVV1    : 15  : Central volume (volume)
TVV2   : 15  : "shallow" Peripheral volume (volume)
TVV3   : 15  : "deep" Peripheral volume (volume)
TVQ    : 5   : Intercompartmental clearance (volume/time)
TVQ2   : 1   : Intercompartmental clearance 2 (volume/time)

$CMT  @annotated
ABS    : Extravascular compartment (mass)
CENT   : Central compartment (mass)
PER    : Peripheral compartment (mass)
PER2   : Peripheral compartment 2 (mass)

$OMEGA @annotated
eKA : 0.09  : eta KA
eCL : 0.09  : eta CL
eV1 : 0.09  : eta V1
eV2 : 0.09  : eta V2
eV3 : 0.09  : eta V3
eQ  : 0.09  : eta Q
eQ2 : 0.09  : eta Q2

$GLOBAL
#define CP (CENT/V1)
#define CST (PER/V2)
#define CDT (PER2/V3)

$MAIN
double KA = TVKA*exp(eKA);
double CL = TVCL*exp(eCL);
double V1 = TVV1*exp(eV1);
double V2 = TVV2*exp(eV2);
double V3 = TVV3*exp(eV3);
double Q = TVQ*exp(eQ);
double Q2 = TVQ2*exp(eQ2);

$ODE
dxdt_ABS    = -KA*ABS;
dxdt_CENT   =  KA*ABS - (CL+Q+Q2)*CP + Q*CST + Q2*CDT;
dxdt_PER    =  Q*CP - Q*CST;
dxdt_PER2   =  Q2*CP - Q2*CDT;

$CAPTURE @annotated
CP : Plasma concentration (mass/volume)
'
pk3 <- mcode("pk3", pk3_code)
```

```{r}
#| warning: false
#| message: false
#| error: false
e <- data.frame(ID=1, amt=300, ii=24, addl=5, cmt=1, evid=1)

pk3 %>%
  ev(e) %>%
  zero_re() %>%
  mrgsim(end = 200, delta=1) %>% 
  plot(CP~time, logy=TRUE)
```

### Nonlinear Pharmacokinetics

Nonlinear PK govern disposition when there is a disproportionate change in tissue or serum concentrations with increasing dose. Often this is due to saturation of one the ADME processes.

| ADME Process | Example                                                                      |
|-------------------|-----------------------------------------------------|
| Absorption   | Saturable transport of p-glycoprotein transporters in the intestine          |
| Distribution | Plasma protein binding; cellular uptake                                      |
| Metabolism   | Saturation of CYP450 enzymes                                                 |
| Elimination  | Saturation of renal filtration, tubular secretion, reabsorption transporters |

#### Michaelis-Menten Kinetics

$$
V = \frac{-dX}{dt}=\frac{V_{max}}{K_m+C}
$$

where $V$ is the velocity of the reaction, $V_{max}$ is the maximum velocity, $C$ is the concentration, and $K_m$ is the concentration at which half the maximal velocity of the reaction is achieved. Notice how the slopes become parallel when the concentration is in the linear elimination phase.

```{r}
#| warning: false
#| message: false
#| error: false
nlcl <- '
$PARAM @annotated
V    : 10  : Central volume (volume)
VMAX :  2  : Maximum reaction velocity (mass/time)
KM   :  5  : Michaelis constant (mass/volume)
    
$CMT  @annotated
CENT   : Central compartment (mass)

$GLOBAL
#define CONC (CENT/V)
#define CL (VMAX/(KM+CONC))*V

$ODE
dxdt_CENT =  - CL*CONC;

$CAPTURE @annotated
CONC : Plasma concentration (mass/volume)
VMAX : Maximum reaction velocity (mass/time)
KM   : Michaelis constant (mass/volume)
CL   : Clearance (L/hr)
'

nlcl <- mcode('nlcl', nlcl)
```

```{r}
#| warning: false
#| message: false
#| error: false
e <- data.frame(ID=1:3, amt=c(50, 100, 300), cmt=1, evid=1)

nlcl %>%
  ev(e) %>%
  mrgsim(end = 48, delta=1) %>% 
  plot(CONC~time, logy=TRUE)
```

### Nonlinear Distribution

**NOT DONE**. Would like to include a cellular compartment to show saturatble distribution into blood cells. The example below demonstrates a case of nonlinear distribution where the drug is taken up by cells in the central compartment via a saturable mechanism.

```{r}
#| warning: false
#| message: false
#| error: false
nld <- '
$PARAM @annotated
VC   : 10  : Central volume (volume)
VP   : 2    : Cell volume (volume)
k10   : 0.3   : Clearance from central compartment (mass/time)
k12   : 0.1   : transfer cent to per (mass/time)
k21   : 0.1   : transfer per to cent (mass/time)
VMAX :  0.3  : Maximum reaction velocity (mass/time)
KM   :  5  : Michaelis constant (mass/volume)
    
$CMT  @annotated
CENT   : Central compartment (mass)
PER    : Peripheral compartment (mass)

$GLOBAL
#define CP (CENT/VC)
#define CT (PER/VP)
#define k12 (VMAX/(KM+CP))

$ODE
dxdt_CENT =  - k10*CENT - k12*CENT + k21*PER;
dxdt_PER = k12*CENT - k21*PER;

$CAPTURE @annotated
CP   : Plasma concentration (mass/volume)
CT   : Cell concentration (mass/volume)
VMAX : Maximum reaction velocity (mass/time)
KM   : Michaelis constant (mass/volume)
'

nld <- mcode('nld', nld)
```

```{r}
#| warning: false
#| message: false
#| error: false
e <- data.frame(ID=1:5, amt=c(100, 200, 500, 1000, 2000), cmt=1, evid=1)

nld %>%
  ev(e) %>%
  mrgsim(end = 24, delta=1) %>% 
  plot(CENT+PER~time, logy=TRUE)
```

## Absorption Models

Drug absorption describes how the drug enters the central plasma and there are many models to describe the complexities of this process. Modeling drug absorption is one of the trickiest aspects of compartmental modeling. Often, there is just not enough data in the absorption phase to fit the model you really want to fit, so you are forced to fit one of the simpler models.

### First-order absorption with bioavailability

First order absorption represents a constant fraction of the available drug in the absorption compartment entering the central compartment per unit time.

```{r}
#| warning: false
#| message: false
#| error: false
first_order_F <- '
$PARAM @annotated
TVKA   : 0.1 : Absorption rate constant (1/hr)
TVF    : 0.7 : typical value bioavailability fraction (%)
TVCL   :  6  : Clearance (volume/time)
TVV1    : 15  : Central volume (volume)

$CMT  @annotated
ABS    : Extravascular compartment (mass)
CENT   : Central compartment (mass)

$OMEGA @annotated
eKA : 0.09  : eta KA
eCL : 0.09  : eta CL
eV1  : 0.09  : eta V

$GLOBAL
#define CP (CENT/V1)

$MAIN
F_ABS = TVF;
double KA = TVKA*exp(eKA);
double CL = TVCL*exp(eCL);
double V1 = TVV1*exp(eV1);

$ODE
dxdt_ABS   = -KA*ABS;
dxdt_CENT  =  KA*ABS - CL*CP;

$SIGMA @annotated
ADD  : 0.1   : additive error

$CAPTURE @annotated
CP : Plasma concentration (mass/volume)
'
  
first_order_F <- mcode('first_order_F', first_order_F)
```

```{r}
#| warning: false
#| message: false
#| error: false
e <- data.frame(ID=1:5, amt=100, ii=24, addl=3, cmt=1, evid=1)

first_order_F %>%
  ev(e) %>%
  mrgsim(end = 100, delta=1) %>% 
  plot(CP~time)
```

### Zero-order absorption

Zero-order absorption describes an absorption process in which a constant amount of drug enters the central compartment from the absorption compartment per unit time.

```{r}
#| warning: false
#| message: false
#| error: false
zero_order_abs <- '
$PARAM @annotated 
CL   : 5   : 
V1   :  20 :
DUR  : 3 : 
 
$CMT CENT

$GLOBAL
#define CP (CENT/V1)

$MAIN
D_CENT = DUR;
 
$ODE
dxdt_CENT = -CL*CP;

$CAPTURE @annotated
CP : plasma concentration
CL : clearance
V1 : volume of distribution
DUR : duration of zero-order absorption
'

zero_order_abs <- mcode('zero_order_abs', zero_order_abs)
```

```{r}
#| warning: false
#| message: false
#| error: false
e <- data.frame(ID=1, amt=200, rate=-2, cmt=1, time=0, evid=1, DUR=5, CL=10)

zero_order_abs %>%
  ev(e) %>%
  mrgsim(end = 24, delta=0.1) %>%
  plot(CP~time, logy=TRUE)
```

### First order absorption with LAG time

This model is often used when there are delays in the absorption profile (e.g., with a delayed release formulation) and the PK data in the absorption phase are sparse. It often improves the model fit, but may be criticized for being un-physiological. I'd agree in most situations, but think this type of model is very reasonable when you have an actual delayed release formulation. Another limitation to this model is that it can cause numerical difficulties with some of the estimation algorithms (FOCE). These ODE solvers really don't like when the value of compartments is 0, and this is exactly the case with this model for time points before the LAG time.

```{r}
#| warning: false
#| message: false
#| error: false
first_order_abs_w_lag <- '
$PARAM @annotated
TVKA   : 0.6 : Absorption rate constant (1/hr)
TVTLAG : 1.5 : typical value lag time (time)
TVCL   :  6  : Clearance (volume/time)
TVV1    : 15  : Central volume (volume)

$CMT  @annotated
ABS    : Extravascular compartment (mass)
CENT   : Central compartment (mass)

$OMEGA @annotated
eKA : 0.09  : eta KA
eTLAG : 0.09 : eta TLAG
eCL : 0.09  : eta CL
eV1  : 0.09  : eta V

$GLOBAL
#define CP (CENT/V1)

$MAIN
double TLAG = TVTLAG*exp(eTLAG);
ALAG_ABS = TLAG;
double KA = TVKA*exp(eKA);
double CL = TVCL*exp(eCL);
double V1 = TVV1*exp(eV1);

$ODE
dxdt_ABS   = -KA*ABS;
dxdt_CENT  =  KA*ABS - CL*CP;

$CAPTURE @annotated
CP : Plasma concentration (mass/volume)
TVTLAG : lag time
TVKA : individual KA
TVCL : individual CL
TVV1 : individual V1 
'
first_order_abs_w_lag <- mcode('first_order_abs_w_lag', first_order_abs_w_lag)
```

```{r}
#| warning: false
#| message: false
#| error: false
e <- data.frame(ID=1:5, amt=100, ii=24, addl=3, cmt=1, evid=1, TVTLAG=3)

first_order_abs_w_lag %>%
  ev(e) %>%
  mrgsim(end = 100, delta=1) %>% 
  plot(CP~time)
```

### Mixed zero- and first- order absorption

Sometimes a combination of absorption processes may be a reasonably hypothesized. The sequential linked zero-order then first-order model can be interpreted in a mechanistic way if the concentration in the gut is initially at the solubility limit for the drug. As drug is absorbed and the concentration falls below the solubility limiT then the process converts from a zero-order input to a first-order input. Note that ALAG1 (lag time of the first order process) must be equal to TK0 (the duration of the zero-order process) plus any lagtime for the zero-order process. This ensures that the first-order process will take over at the end of the zero-order input.

**NOT DONE.**

```{r}
#| warning: false
#| message: false
#| error: false
mixed_abs <- '
$PARAM @annotated
CL :  1  : clearance
VC : 20  : central volume
KA : 1.1 : first order absorption rate
FRAC : 0.2 : fraction
DUR : 3 : duration of zero-order absorption
 
$CMT @annotated
ABS1 : absorption compartment 1
ABS2 : absorption compartment 2
CENT : Central compartment

$MAIN
F_ABS = 1-FRAC;
F_CENT = FRAC;
D_CENT = DUR;

$GLOBAL
#define CP (CENT/VC)
 
$ODE
dxdt_ABS = -KA*ABS;
dxdt_CENT = KA*ABS - (CL/VC)*CENT;

$CAPTURE @annotated
CP : Plasma concentration
'

mixed_abs <- mcode('mixed_abs', mixed_abs)
```

```{r}
#| warning: false
#| message: false
#| error: false
```

### Weibull Absorption

The Weibull distribution may be used to model time-varying absorption rates. There are many ways to parameterize a weibull model.

$$
KA_{(t)}=\frac{\beta}{\alpha}(\frac{time}{\alpha})^{\beta-1}
$$

```{r}
#| warning: false
#| message: false
#| error: false
weibull_abs <- '
$PARAM @annotated 
CL   : 5   : clearance
VC   :  20 : volume of distribution
BETA : 3   : beta
ALPHA: 2   : alpha
 
$CMT ABS CENT

$MAIN
double KA = (BETA/ALPHA)*pow((TIME/ALPHA), (BETA-1));
double CP = CENT/VC;
 
$ODE
dxdt_ABS = -KA*ABS;
dxdt_CENT = KA*ABS - (CL/VC)*CENT;

$CAPTURE @annotated
CP : plasma concentration 
KA : absorption rate constant
BETA : weibull parameter
ALPHA : weibull parameter
'

weibull_abs <- mcode('weibull_abs', weibull_abs)
```

```{r}
#| warning: false
#| message: false
#| error: false
weibull_abs %>%
  ev(amt=100) %>%
  mrgsim(end=20, delta=0.1) %>%
  plot(KA+CP~time)
```

### Transit Absorption

```{r}
#| warning: false
#| message: false
#| error: false
transit_abs <- '
$PARAM @annotated
KA   : 0.6 : Absorption rate constant (1/hr)
CL   :  6  : Clearance (volume/time)
V1    : 15  : Central volume (volume)
MTT   : 20   : mean transit time
  
$CMT  @annotated
ABS    : Extravascular compartment (mass)
TRAN1  : Transit compartment 1
TRAN2  : Transit compartment 2
TRAN3  : Transit compartment 3
CENT   : Central compartment (mass)

$GLOBAL
#define CP (CENT/V1)

$MAIN
double ktr = 4/MTT;
double k12 = ktr;
double k23 = ktr;
double k34 = ktr;
double k45 = ktr;
double k10 = CL/V1;

$ODE
dxdt_ABS   = -k12*ABS;
dxdt_TRAN1 = k12*ABS - k23*TRAN1;
dxdt_TRAN2 = k23*TRAN1 - k34*TRAN2;
dxdt_TRAN3 = k34*TRAN2 - k45*TRAN3;
dxdt_CENT  =  k45*TRAN3 - k10*CENT;

$CAPTURE @annotated
CP : Plasma concentration (mass/volume)
MTT : mean transit time
'
  
transit_abs <- mcode('transit_abs', transit_abs)
```

```{r}
#| warning: false
#| message: false
#| error: false
transit_abs %>%
  ev(amt=100) %>%
  mrgsim(end=30, delta=1) %>%
  plot(TRAN3+CENT+TRAN1+TRAN2~time)
```

### Enterohepatic Recycling

TBD.

### References
